\sys: Scalable, Property-Guided\\ Program Regeneration

### 2025-07-31 docs/plos/draft.tex
Fix typos
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

Modern software systems evolve constantly.
Developers refactor legacy code~\cite{Fowler99,Mens04,facebook2010redesigns,dropbox2014syncengine},
adapt libraries to changing APIs~\cite{dig2005role,kula2017empiricalstudyimpactrefactoring},
translate components across languages~\cite{manzoor_cli_python,gaultier_rewrite_cpp},
% and patch security vulnerabilities~\cite{ikegami2022userefactoringsecurityvulnerability,schneier2013security_vulnerabilities}.
These transformations, are often brittle, time-consuming, and require significant expertise.

Recent advances in natural-language processing, particularly large language
models (LLMs) have opened new possibilities for automating program
transformation and synthesis tasks,
making them attractive tools for software
evolution, albeit with caveats.
This flexibility has positioned them as attractive tools for software evolution
compared to traditional program synthesis techniques, which often struggle
to scale to real-world settings or across languages and require significant developer involvement~\cite{reynolds2019syguscomp,leino2016dafny,wu2023programming,dynamoth2016,cambronero2019active}.
LLMs can translate code across languages~\cite{ou2025enhancingllmbasedcodetranslation},
refactor complex modules~\cite{ziftci2025migrating},
and even generate entire components from scratch~\cite{huynh2025largelanguagemodelscode}.
However, surface-level patterns in the input shape their outputs~\cite{yang2025evaluatinggeneralizationcapabilitieslarge},
they struggle with negative constraints~\cite{hwang2024thinkpinkelephant,jiang2024llmsdreamelephantswhen},
often overfit to spurious input cues~\cite{xu2023llmfoolitselfpromptbased, wu2023deceptpromptexploitingllmdrivencode},
and lack mechanisms for enforcing precise behavioral properties on the generated code~\cite{roh2025breakthechainreasoningfailuresllms}.
As a result, developers must prompt and steer them through ad-hoc heuristics, manual trial-and-error, or high-quality (and thus hard to create) examples~\cite{khattab2023dspy,brown2020languagemodelsfewshotlearners},
leading to workflows that are difficult to audit, integrate, or systematically reason about.
Bridging this gap calls for an that allows developers to specify explicit semantic constraints and transformation goals, and to reason about their satisfaction in a principled way.

@@

Consider a complex application, like the one
sketched in \cref{fig:ex-large-app}.
A single codebase includes many needs for transformations.
For example, a developer going through older parts of the codebase written by 
a previous team might realize that many functions are susceptible to buffer-overflow attacks.
Then, thinking about a recently incorporated JavaScript scripting component 
that allows players to make custom extensions to key parts of client-side game logic, 
suggests that perhaps code that deals with user input should all be moved inside 
the sandboxed runtime.
Furthermore, the development team would like to identify functions that could 
be broken-down further to assist with performance profiling.

\heading{The state of the art and practise}
Currently, a developer has access to a myriad of different approaches 
to attack the above problems, but all come with shortcomings.
Code quality tools~\cite{avgustinov2016ql, Fowler99,Mens04,Myers16} and specialized transformation systems might be able to detect 
code-smells or certain classes of vulnerabilities but are often very highly 
tied to their target language.
As the programming-language landscape is becoming increasingly diverse,
and developers are building systems where multiple different languages are runtimes co-exist,
these limitations cannot be ignored.
Program synthesis could be an attractive avenue that allows 
the developer to 
The ever-present limitation of traditional program synthesis techniques~\cite{jha2010oracle, raza2018disjunctive, singh2016blinkfill,wu2023programming, harp:ccs:2021, yaghmazadeh2018automated} is their 
inability to scale beyond narrowly-defined computational domains.
This scalability barrier is what statistical techniques like LLMs 
have promised to solve.
An LLM could in principle be tasked with performing all the aforementioned transformations.
However, they come with no shortage of well-studied shortcomings~(\cref{sec:intro}).
There have been systems that provide ranging amounts of control of LLM-generated 
programs~\cite{huang2023anplnaturalprogramminginteractive,li2025mccoderstreamliningmotioncontrol,barua2025pygencollaborativehumanaiapproach, tihanyi2025new},
but there is a large gap of a unifying framework that identifies 
key parts of the problems and applies LLMs in a way that underlines their strengths,
end-users to steer their output, and---more importantly---be aware when 
their outputs are \emph{wrong}.

@@

\begin{figure}[t]
\centering
  % https://docs.google.com/drawings/d/1zMcTurJ8twCG4BrkSUAJersbNSl5UQkogHqqg88vG9M/edit
  \includegraphics[width=.9\columnwidth]{figs/kvasir_application.pdf}
  \caption{\textbf{Applying \sys to a complex game application.}
  The high level of modularity present in modern applications allows \sys to be applied effectively inside a codebase.
  In the above graph, edges signify logical dependencies between components 
  and shaded vertices signify parts of the codebase \sys is applied to~(\cref{sec:example}).
  Dotted edges signify ommited layers of dependencies.
  }
  \label{fig:ex-large-app}
\end{figure}

@@

\begin{wrapfigure}[4]{r}{.53\columnwidth}
\begin{minted}{prolog}
fsignt(p_,S):-fsignt(p,S).
graph(p_,I,O):-graph(p,I,O),
               crash(p,I).
:-crash(p_,I),graph(p_,I,_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \ttt{P\_} to exhibit the same
input-output behavior on representative inputs that would cause a buffer overflow in the original program \ttt{P}.
but not on the regenerated program \ttt{P\_}.
\sys first arrives the set of properties to extract from the
original program, given the query:
\{\prop{input(P, I)}, \prop{graph(P, I, O)}, \prop{fsignt(P, S)}\}, where
\prop{input(P, I)} means that \sys will use extract a set of inputs from the original program
by prompting an LLM with the program's source code and asking it to generate a set of inputs that
would cause a buffer overflow in the original program;
\prop{graph(P, I, O)} extracts the client-observable input-output behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \prop{fsignt(P, S)} extracts each function's signature using a language-aware parser.
After 30.4 seconds and two attempts, \sys synthesizes the following program:
\begin{minted}[]{js}
#define MAX_C 1024
char *Args(int argc, char **argv) {
static char args[MAX_C];
int i; size_t remaining = MAX_C - 1;
args[0] = 0;
for (i = 1; i < argc; i++) {
  strncat(args, argv[i], remaining);
  remaining = MAX_C - strlen(args) - 1;
  if (i != argc - 1 && remaining > 0) {
    strncat(args, " ", remaining);
    remaining = MAX_C - strlen(args) - 1;
  } }
  return args; }
\end{minted}

@@

\heading{Prototype implementation}
The \sys prototype implemetation uses Python v3.13,
the \ttt{pluggy} system~\cite{pluggy} to define plugin hooks,
compiles queries to the Clingo language~\cite{DBLP:journals/corr/GebserKKS14} and uses the \ttt{clasp}~\cite{gebser2007clasp}
solver to arrive at a stable model which is converted to synthesis plans. 
A total of seven plugins have been implemented (\ttt{graph}, \ttt{tr}, \ttt{len}, \ttt{sql\_trc}, \ttt{call}, \ttt{crash}, \ttt{fsignt}).
Each plugin ranges from 9 to 51 lines-of-code,
with the entire codebase being 605 lines-of-code.

\heading{A benchmark suite for program regeneration}
For research on program regeneration to flourish, it needs a common benchmark
suite that captures the diversity and difficulty of real-world transformations.
Today, there is no equivalent of SPEC for performance or Defects4J for bug
fixing---researchers and practioners rely on ad hoc examples that are hard to compare and often
oversimplified.
This lack of a shared baseline obscures progress. 
A useful set of benchmarks should include
programs that
(1) are real and non-trivial,
(2) are written in a variety of languages, 
(3) cover a wide range of computational domains.
Each benchmark should include an original program, 
the same program that undergoes the transformation described in natural-language 
in an accompanying file, and a set of tests that verify that the transformation was succesful.
Following the lead of benchmarking efforts~\cite{SPEC_CPU2017, dacapo2006, koala:atc:2025}, the suite should be made open-source, community 
first, and an established technical steering commitee should guide extensions and modifications towards maintaining the suite's relevance.

@@

The so-far 
described subsystems~(\cref{sec:design}) can be combined to solve all transformation
goals in the snippets shown so far~(\cref{sec:example}).
Scaling these techniques further, however, poses more challegnes and uncovers implications.

>>>>

Modern software systems evolve constantly.
Developers refactor legacy code~\cite{Fowler99,Mens04,facebook2010redesigns,dropbox2014syncengine},
adapt libraries to changing APIs~\cite{dig2005role,kula2017empiricalstudyimpactrefactoring},
translate components across languages~\cite{manzoor_cli_python,gaultier_rewrite_cpp},
% and patch security vulnerabilities~\cite{ikegami2022userefactoringsecurityvulnerability,schneier2013security_vulnerabilities}.
These transformations are often brittle, time-consuming, and require significant expertise.

Recent advances in natural-language processing, particularly large language
models (LLMs) have opened new possibilities for automating program
transformation and synthesis tasks,
making them attractive tools for software
evolution, albeit with caveats.
This flexibility has positioned them as attractive tools for software evolution
compared to traditional program synthesis techniques, which often struggle
to scale to real-world settings or across languages and require significant developer involvement~\cite{reynolds2019syguscomp,leino2016dafny,wu2023programming,dynamoth2016,cambronero2019active}.
LLMs can translate code across languages~\cite{ou2025enhancingllmbasedcodetranslation},
refactor complex modules~\cite{ziftci2025migrating},
and even generate entire components from scratch~\cite{huynh2025largelanguagemodelscode}.
However, surface-level patterns in the input shape their outputs~\cite{yang2025evaluatinggeneralizationcapabilitieslarge},
they struggle with negative constraints~\cite{hwang2024thinkpinkelephant,jiang2024llmsdreamelephantswhen},
often overfit to spurious input cues~\cite{xu2023llmfoolitselfpromptbased, wu2023deceptpromptexploitingllmdrivencode},
and lack mechanisms for enforcing precise behavioral properties on the generated code~\cite{roh2025breakthechainreasoningfailuresllms}.
As a result, developers must prompt and steer them through ad-hoc heuristics, manual trial-and-error, or high-quality (and thus hard to create) examples~\cite{khattab2023dspy,brown2020languagemodelsfewshotlearners},
leading to workflows that are difficult to audit, integrate, or systematically reason about.
Bridging this gap calls for an approach that allows developers to specify explicit semantic constraints and transformation goals, and to reason about their satisfaction in a principled way.

@@

Consider a complex application, like the one
sketched in \cref{fig:ex-large-app}.
A single codebase includes many needs for transformations.
For example, a developer going through older parts of the codebase written by 
a previous team might realize that many functions are susceptible to buffer-overflow attacks.
Then, thinking about a recently incorporated JavaScript scripting component 
that allows players to make custom extensions to key parts of client-side game logic, 
suggests that perhaps code that deals with user input should all be moved inside 
the sandboxed runtime.
Furthermore, the development team would like to identify functions that could 
be broken down further to assist with performance profiling.

\heading{The state of the art and practise}
Currently, a developer has access to a myriad of different approaches 
to attack the above problems, but all come with shortcomings.
Code quality tools~\cite{avgustinov2016ql, Fowler99,Mens04,Myers16} and specialized transformation systems might be able to detect 
code-smells or certain classes of vulnerabilities but are often very highly 
tied to their target language.
As the programming-language landscape is becoming increasingly diverse,
and developers are building systems where multiple different languages are runtimes coexist,
these limitations cannot be ignored.
Program synthesis could be an attractive avenue that allows 
the developer to 
The ever-present limitation of traditional program synthesis techniques~\cite{jha2010oracle, raza2018disjunctive, singh2016blinkfill,wu2023programming, harp:ccs:2021, yaghmazadeh2018automated} is their 
inability to scale beyond narrowly-defined computational domains.
This scalability barrier is what statistical techniques like LLMs 
have promised to solve.
An LLM could in principle be tasked with performing all the aforementioned transformations.
However, they come with no shortage of well-studied shortcomings~(\cref{sec:intro}).
There have been systems that provide ranging amounts of control of LLM-generated 
programs~\cite{huang2023anplnaturalprogramminginteractive,li2025mccoderstreamliningmotioncontrol,barua2025pygencollaborativehumanaiapproach, tihanyi2025new},
but there is a large gap of a unifying framework that identifies 
key parts of the problems and applies LLMs in a way that underlines their strengths,
end-users to steer their output, and---more importantly---be aware when 
their outputs are \emph{wrong}.

@@

\begin{figure}[t]
\centering
  % https://docs.google.com/drawings/d/1zMcTurJ8twCG4BrkSUAJersbNSl5UQkogHqqg88vG9M/edit
  \includegraphics[width=.9\columnwidth]{figs/kvasir_application.pdf}
  \caption{\textbf{Applying \sys to a complex game application.}
  The high level of modularity present in modern applications allows \sys to be applied effectively inside a codebase.
  In the above graph, edges signify logical dependencies between components 
  and shaded vertices signify parts of the codebase \sys is applied to~(\cref{sec:example}).
  Dotted edges signify omitted layers of dependencies.
  }
  \label{fig:ex-large-app}
\end{figure}

@@

\begin{wrapfigure}[4]{r}{.53\columnwidth}
\begin{minted}{prolog}
fsignt(p_,S):-fsignt(p,S).
graph(p_,I,O):-graph(p,I,O),
               crash(p,I).
:-crash(p_,I),graph(p_,I,_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \ttt{P\_} to exhibit the same
input-output behavior on representative inputs that would cause a buffer overflow in the original program \ttt{P}
but not on the regenerated program \ttt{P\_}.
\sys first arrives the set of properties to extract from the
original program, given the query:
\{\prop{input(P, I)}, \prop{graph(P, I, O)}, \prop{fsignt(P, S)}\}, where
\prop{input(P, I)} means that \sys will use extract a set of inputs from the original program
by prompting an LLM with the program's source code and asking it to generate a set of inputs that
would cause a buffer overflow in the original program;
\prop{graph(P, I, O)} extracts the client-observable input-output behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \prop{fsignt(P, S)} extracts each function's signature using a language-aware parser.
After 30.4 seconds and two attempts \sys synthesizes the following program:
\begin{minted}[]{js}
#define MAX_C 1024
char *Args(int argc, char **argv) {
static char args[MAX_C];
int i; size_t remaining = MAX_C - 1;
args[0] = 0;
for (i = 1; i < argc; i++) {
  strncat(args, argv[i], remaining);
  remaining = MAX_C - strlen(args) - 1;
  if (i != argc - 1 && remaining > 0) {
    strncat(args, " ", remaining);
    remaining = MAX_C - strlen(args) - 1;
  } }
  return args; }
\end{minted}

@@

\heading{Prototype implementation}
The \sys prototype implementation uses Python v3.13,
the \ttt{pluggy} system~\cite{pluggy} to define plugin hooks,
compiles queries to the Clingo language~\cite{DBLP:journals/corr/GebserKKS14} and uses the \ttt{clasp}~\cite{gebser2007clasp}
solver to arrive at a stable model which is converted to synthesis plans. 
A total of seven plugins have been implemented (\ttt{graph}, \ttt{tr}, \ttt{len}, \ttt{sql\_trc}, \ttt{call}, \ttt{crash}, \ttt{fsignt}).
Each plugin ranges from 9 to 51 lines-of-code,
with the entire codebase being 605 lines-of-code.

\heading{A benchmark suite for program regeneration}
For research on program regeneration to flourish, it needs a common benchmark
suite that captures the diversity and difficulty of real-world transformations.
Today, there is no equivalent of SPEC for performance or Defects4J for bug
fixing---researchers and practitioners rely on ad hoc examples that are hard to compare and often
oversimplified.
This lack of a shared baseline obscures progress. 
A useful set of benchmarks should include
programs that
(1) are real and non-trivial,
(2) are written in a variety of languages, 
(3) cover a wide range of computational domains.
Each benchmark should include an original program, 
the same program that undergoes the transformation described in natural-language 
in an accompanying file, and a set of tests that verify that the transformation was successful.
Following the lead of benchmarking efforts~\cite{SPEC_CPU2017, dacapo2006, koala:atc:2025}, the suite should be made open-source, community 
first, and an established technical steering committee should guide extensions and modifications towards maintaining the suite's relevance.

@@

The so-far 
described subsystems~(\cref{sec:design}) can be combined to solve all transformation
goals in the snippets shown so far~(\cref{sec:example}).
Scaling these techniques further, however, poses more challenges and uncovers implications.
